A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

Given a signature $\Sig$, we write $\thD$ to refer to the \emph{theory} of (co-)datatypes,
which in addition to $\Sig$ defines a class of $\Sig$-interpretations $\mathbf{I}$,
namely the ones that satisfy the axioms mentioned in this section.
In particular, this means that each interpretation in $\mathbf{I}$ has a fixed interpretation
for constructor terms and correctly-applied selector terms, but may differ on its interpretation for
wrongly-applied selector terms and variables.
A formula $\varphi$ is $\thD$-satisfiable if there is an interpretation in $\mathbf{I}$ that satisfies $\varphi$,
and $\thD$-unsatisfiable otherwise.
\rem{more?}

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Unfold} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{example}

@@

\begin{example}
Let $\const{C} : $ $\tau_1 \times \tau_2 \rightarrow \tau_2$ be a constructor,
and let $\Ec = \{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
Constructing $\Val$ as described above yields
\[\begin{array}{rcrlrl}
\Val \ec{x} & \,=\, &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, & \mu {\VAREC{\const{C}(y, x)}}. & \const{C}( {\Varec{y}},\, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\VAREC{\const{C}(y,x)}}. & \const{C}( {\Varec{y}},\, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, {\VAREC{\const{C}(y,x)}} ) )
\end{array}\]
Since $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are $\alpha$-equivalent,
%for the renaming
%$\{ {\Varec{x}} \mapsto {\VAREC{\const{C}(y,x)}}, {\VAREC{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
the calculus can infer $x \teq \const{C}(y,x)$.
%Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
%are equivalent for any value of ${\Varec{y}}$.
\xend
\end{example}

@@

\paragraph{Phase 3: Branching.}
In the third phase,
the \rn{Singleton} rule handles the case when there are two terms from $\tEc$
having a co-recursive singleton type, in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type,
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$),
which is enforced by the rule \rn{Split}.

>>>>

Given a signature $\Sig$, we write $\thD$ to refer to the \emph{theory} of (co)datatypes,
which in addition to $\Sig$ defines a class of $\Sig$-interpretations $\mathbf{I}$,
namely the ones that satisfy the axioms mentioned in this section.
In particular, this means that each interpretation in $\mathbf{I}$ has a fixed interpretation
for constructor terms and correctly applied selector terms, but may differ on its interpretation for
wrongly applied selector terms and variables.
A formula $\varphi$ is $\thD$-satisfiable if there exists an interpretation in $\mathbf{I}$ that satisfies $\varphi$;
otherwise, it is $\thD$-unsatisfiable.
\rem{more?}

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Unfold} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initially:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{example}

@@

\begin{example}
Let $\const{C} : $ $\tau \times \delta \rightarrow \delta$ be a constructor,
and let $\Ec = \{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
Constructing $\Val$ as described above yields
\[\begin{array}{rcrlrl}
\Val \ec{x} & \,=\, &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, & \mu {\VAREC{\const{C}(y, x)}}. & \const{C}( {\Varec{y}},\, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\VAREC{\const{C}(y,x)}}. & \const{C}( {\Varec{y}},\, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, {\VAREC{\const{C}(y,x)}} ) )
\end{array}\]
Since $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are $\alpha$-equivalent,
%for the renaming
%$\{ {\Varec{x}} \mapsto {\VAREC{\const{C}(y,x)}}, {\VAREC{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
the calculus can infer $x \teq \const{C}(y,x)$.
%Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
%are equivalent for any value of ${\Varec{y}}$.
\xend
\end{example}

@@

\paragraph{Phase 3: Branching.}
In the third phase,
the \rn{Singleton} rule handles the case when there are two terms from $\tEc$
having a corecursive singleton type, in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type,
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$),
which is enforced by the rule \rn{Split}.
